---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BBNI

<!-- badges: start -->
[![R-CMD-check](https://github.com/anson-li8/BBNI/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/anson-li8/BBNI/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

BBNI implements a full Bayesian approach for inferring Boolean genetic networks from noisy gene expression data, based on the method in [Han et al. (2014)](https://doi.org/10.1371/journal.pone.0115806). Unlike methods that return a single best-fit topology (REVEAL & BFE, which are both found in [BoolNet](https://cran.r-project.org/package=BoolNet)), BBNI uses MCMC to sample the full joint posterior over network topologies and Boolean transition functions, explicitly accounting for biological noise and returning posterior edge probabilities rather than one point estimate.

## Installation

BBNI is not yet on CRAN. Install the development version from GitHub:

``` r
# install.packages("devtools")
devtools::install_github("anson-li8/BBNI")
```

## Example

A minimal check that the package loads and runs:

```{r example}
library(BBNI)

set.seed(1)
true_network <- GenerateNetwork(num.node = 5)
dummy_data <- GenerateSample(
  trans_matrix = true_network,
  num.node = 5,
  SampleSize = 20,
  para = rep(0.5, 5),
  error = matrix(0, nrow = 5, ncol = 20)
)
dummy_data
```

For a complete walkthrough running the MCMC sampler, checking convergence, and evaluating network recovery, please see the [Introduction to BBNI](https://anson-li8.github.io/BBNI/articles/Introduction_to_BBNI.html) vignette.

## Citation

```{r, echo = FALSE, comment = ""}
print(citation("BBNI"), bibtex = FALSE)
```

## License

MIT
