---
title: "Introduction to BBNI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BBNI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `BBNI` (Boolean Genetic Network Inference) package provides a Bayesian approach for inferring both the topology and the Boolean logic transition functions of genetic networks from time-series expression data. 

As described in Han et al. (2014), biological systems often exhibit noise and uncertainty. `BBNI` uses a Metropolis-within-Gibbs Markov chain Monte Carlo (MCMC) algorithm to sample from the joint posterior distribution of valid Directed Acyclic Graph (DAG) topologies and their corresponding Boolean rules, specifically accounting for biological noise.

## Setup and Data Simulation

First, load the package:

```{r setup}
library(BBNI)
```

To test the MCMC sampler, we will use the package's built-in simulation functions to generate a artificial true network topology and its corresponding noisy time-series data. This allows us to verify if the algorithm can recover the true topology.

```{r simulate-data}
set.seed(235)
num_nodes <- 10
sample_size <- 50

# 1. Generate a true network topology (T) and Boolean rules (F)
true_network <- GenerateNetwork(num.node = num_nodes)

# 2. Setup parameters for data generation
# Baseline success probabilities for root nodes
root_probs <- rep(0.5, num_nodes) 

# Generate a binary noise matrix (e.g., 5% natural error rate)
error_rate <- 0.05
error_matrix <- matrix(rbinom(num_nodes * sample_size, 1, error_rate), 
                       nrow = num_nodes, 
                       ncol = sample_size)

# 3. Simulate the time-series observational data (G)
dummy_data <- GenerateSample(trans_matrix = true_network,
                             num.node = num_nodes,
                             SampleSize = sample_size,
                             para = root_probs,
                             error = error_matrix)
```

## Running the MCMC Sampler

Now we apply the `run_bbni()` algorithm to our synthetic data to see if it can infer the underlying network. We must define a few hyperparameters, including the Beta priors ($\alpha, \beta$) for the root node probabilities and the global noise parameter $e$.

```{r run-model, results='hide'}
# Define hyperparameters
num_update <- 100       # Number of MCMC iterations (kept low for tutorial speed)
penalty <- 0.1          # Structural prior penalty for network complexity
prop_ratio <- 0.1       # Proposal threshold

# Set Beta priors: 
# The last row represents the global noise parameter 'e' (needs beta1 < beta2)
prior_para <- matrix(3, nrow = num_nodes + 1, ncol = 2)
prior_para[num_nodes + 1, 1] <- 2    
prior_para[num_nodes + 1, 2] <- 100  

# Execute the sampler
mcmc_results <- run_bbni(GeneData = dummy_data, 
                         num.node = num_nodes, 
                         SampleSize = sample_size, 
                         prior_para = prior_para, 
                         num_update = num_update, 
                         penalty = penalty, 
                         prop.ratio = prop_ratio)
```

## Analyzing the Output

The `run_bbni()` function returns a list outlining the MCMC chain trajectory. We can now see the final sampled transition matrix and compare the log-posterior scores.

```{r view-results}
# View the log-posterior scores of the final 5 iterations
tail(mcmc_results$log_posterior, 5)

# The final sampled network topology and boolean functions
final_network <- tail(mcmc_results$networks, 1)[[1]]
```